---
title: "Machine Learning Project"
author: "Peter Leap"
date: "April 8, 2015"
output: html_document
---
###Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement? a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, **your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.** They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

###The Question
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. 


Load Packages:
```{r}
require(dplyr, warn.conflicts = FALSE, quietly = TRUE)
require(rpart, warn.conflicts = FALSE, quietly = TRUE)
require(rpart.plot, warn.conflicts = FALSE, quietly = TRUE)
require(rattle, warn.conflicts = FALSE, quietly = TRUE)
require(caret, warn.conflicts = FALSE, quietly = TRUE)
require(randomForest, warn.conflicts = FALSE, quietly = TRUE)
```

###Load data
  * Load from provided data files for train & test
  * Select accelerometers on the belt, forearm, arm and dumbell
  * Keep the classe for training

```{r}
data_train<-read.csv("data/pml-training.csv")
data_test<-read.csv("data/pml-testing.csv")
```

###Analysis
I partitioned the data set into training & testing, selected designated predictors

   * Used the training data provided to partion my own training and testing sets for cross validation
   * Based on the Background, selected only the predictors that are accelerometers.
   * Created a train predictor/feature for use with the radomForest call later
   
```{r}
inTrain <- createDataPartition(y=data_train$classe,
                               p=0.7, list = FALSE)
train <- data_train[inTrain,]
test <- data_train[-inTrain,]
train<- train %>%
  select(starts_with("accel_"), classe)
feature<- train$classe
predictors<- train %>%
  select(starts_with("accel_"))
test<- test %>%
  select(starts_with("accel_"), classe)

```

I used the training sub-set to train the ML algorithms using train() & randomForest() - was much faster than train for df

  * modFit1  is rpart
  * modFit2 is a rf model
  
Plotted the tree for modFit1 and table of importance for modFit2  
```{r}
modFit1 <- train(classe ~ ., method="rpart", data=train)
modFit2<-randomForest(predictors, feature)
fancyRpartPlot(modFit1$finalModel)
varImp(modFit2)
```

I then computed the in sample accuracy using predict/confusionMatrix against the training sub-set
```{r}
confusionMatrix(train$classe, predict(modFit1, train))$overall[1]
confusionMatrix(train$classe, predict(modFit2, train))$overall[1]
```

I selected an modFit2 (random forest) for cross validation on the basis of the in sample accuracies

I used the trained model from that algorithm to predict the outcome for the test subset
```{r}
predictions <- predict(modFit2, test)
```

###Cross Validation/Expected Out of Sample Error
Then used the predicted outcome for the test sub-set with confusionMatrix to estimate the out of sample accuracy.  I expect a 95% accuracy rate.
```{r}
confusionMatrix(test$classe, predictions)$table
confusionMatrix(test$classe, predictions)$overall[1]
```

###Reasons for choices made
I selected the random forest because of the higher accuracies and the good cross validation results I got.

###Model applied to the original test data:
When I submitted my predictions, 19 out of 20 were correct.  (95%!)
```{r}
data_test<- data_test %>%
  select(starts_with("accel_"))
answers<-predict(modFit2, data_test)
answers
```
